---
title: "R Notebook"
output: html_notebook
---



```{r}
# R script accompanying quasi-concave, concave etc note

if(!require(rgl)) install.packages('rgl')
require(rgl)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.
```{r}
# setup data
nx <- 50	# points into x direction
ny <- 50	# points into y

x <- seq(0,1,length=50)
y <- seq(0,1,length=50)

df <- expand.grid(x=x,y=y)
```

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

```{r}
# DGPs
cobb.douglas <- function(x,a,b) return( x[1]^a * x[2]^b )
```

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.



```{r}
# plotting data
cd1 <- apply(df,1,cobb.douglas,a=0.25,b=0.25)
cd2 <- apply(df,1,cobb.douglas,a=2,b=2)
```

```{r}
# put in a matrix
cd.heights1 <- matrix(cd1,nrow=nx,ncol=ny)
cd.heights2 <- matrix(cd2,nrow=nx,ncol=ny)
```

```{r}
# concave cobb douglas
open3d()
surface3d(x,y,cd.heights1,color="red",alpha=0.8)
axes3d(labels=FALSE,tick=FALSE)
title3d("concave cobb douglas")
```

